---
title: "Box-Jenkins Forecasting: SARIMA Modeling of Pharmaceutical Sales"
subtitle: "Evaluating Seasonal Demand and Model Ensembles for Antihistamine Sales (R06)"
author: "Renata Halenka 20201528; João Neto 20201566"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
    highlight: kate
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction & Problem Framing

This project applies the Box-Jenkins methodology to analyze and forecast
the sales volume of pharmaceutical products, specifically focusing on
the R06 category (Antihistamines for systemic use).\
The dataset consists of Pharmacy Point-of-Sale transactions, which have
been aggregated to a monthly frequency.

Because antihistamine demand is heavily driven by real-world biological
cycles, this series is expected to exhibit strong, predictable seasonal
fluctuations.

Accurate forecasting on this localized scale holds significant business
value for individual pharmacies. It allows them to anticipate peak
demand, optimize inventory procurement, and prevent stock-outs.

By validating a parsimonious ARIMA model, this analysis aims to
determine whether short-term antihistamine demand can be reliably
predicted using only its own historical autoregressive and seasonal
memory.

# Exploratory Data Analysis & Preprocessing

```{r data-prep, message=FALSE, warning=FALSE, echo=FALSE}
library(fpp3)
library(lubridate)
library(dplyr)
library(knitr)
library(urca)
library(fable)
library(kableExtra)

meds <- read.csv("salesdaily.csv", sep = ",")

anti_histam <- meds %>%
  mutate(Mes = yearmonth(mdy(datum))) %>% 
  filter(Mes < yearmonth("2019 Oct")) %>%  ##remove incomplete month
  group_by(Mes) %>% 
  summarise(Total_Sales = sum(as.numeric(R06), na.rm = TRUE)) %>% 
  as_tsibble(index = Mes)

 
cutoff <- max(anti_histam$Mes) - 16
anti_histam_train <- anti_histam %>% filter(Mes <= cutoff)
anti_histam_test  <- anti_histam %>% filter(Mes > cutoff)

```

## Visual Inspection and STL Decomposition

```{r stl-decomp, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}

anti_histam_train %>%
  model(STL(Total_Sales ~ trend(window = 21) + season(window = "periodic"))) %>%
  components() %>%
  autoplot() +
  labs(title = "STL Decomposition of Antihistamine Sales",
       subtitle = "Isolating the rigid seasonal allergy cycle from trend and remainder")

```

Visual inspection of the raw antihistamine sales reveals a clear
12-month seasonal pattern. Furthermore, the amplitude of the seasonal
fluctuations is not constant; the variance appears to increase with the
level of the series. Because the variance is non-constant, a variance
stabilization transformation is formally required before proceeding to
stationarity tests.

## Variance Stabilization Evaluation

```{r variance-stabilization, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}
lambda_val <- anti_histam_train %>%
  features(Total_Sales, features = guerrero) %>%
  pull(lambda_guerrero)


anti_histam_train <- anti_histam_train %>%
  mutate(Transformed_Sales = box_cox(Total_Sales, lambda_val))
anti_histam_test <- anti_histam_test %>%
  mutate(Transformed_Sales = box_cox(Total_Sales, lambda_val))
anti_histam <- anti_histam %>%
  mutate(Transformed_Sales = box_cox(Total_Sales, lambda_val))

data.frame(
  Method = "Guerrero Optimization",
  Optimal_Lambda = round(lambda_val, 3),
  Transformation = ifelse(abs(lambda_val - 1) > 0.1,
                          "Box-Cox Required",
                          "No Transformation Needed")
) %>%
  kable(caption = "Variance Stabilization Diagnostic (Training Set)", align = "c")


```

During our initial visual inspection, the amplitude of the seasonal
sales fluctuations appeared to increase alongside the overall level of
the series, suggesting non-constant variance. To formally test this, we
applied the Guerrero method to the training data. The calculation
yielded an optimal Box-Cox parameter (λ) of `r round(lambda_val, 3)`, which is
significantly distinct of 1.0.

This mathematical result confirms our visual hypothesis: the variance is
non-constant. Therefore, the Box-Cox transformation is strictly required
to stabilize the variance across the series, satisfying the prerequisite
assumptions for formal Box-Jenkins estimation.

## Stationarity Diagnostics and Differencing

```{r unit-root-tests, echo=FALSE, message=FALSE, warning=FALSE}



run_gts_adf <- function(series, type = "trend", max_lags = 16) {
  optimal_p <- 0
  for (p in max_lags:1) {
    test <- ur.df(series, type = type, lags = p)
    coefs <- test@testreg$coefficients
    if (coefs[nrow(coefs), 4] < 0.10) {
      optimal_p <- p
      break
    }
  }
  ur.df(series, type = type, lags = optimal_p)

}


adf_raw <- run_gts_adf(anti_histam_train$Transformed_Sales,
                        type = "trend", max_lags = 16)



anti_histam_train <- anti_histam_train %>%
  mutate(Stationary_Sales = difference(Transformed_Sales, 16))



t_test_diff <- t.test(na.omit(anti_histam_train$Stationary_Sales))



adf_diff <- run_gts_adf(na.omit(anti_histam_train$Stationary_Sales),
                         type = "none", max_lags = 16)


adf_summary <- data.frame(
  Data_State       = c("Box-Cox Transformed", "Box-Cox + Seasonal Diff (D=1)"),
  ADF_Type         = c("trend", "none"),
  Optimal_Lags     = c(length(adf_raw@testreg$coefficients[,1]) - 3,
                       length(adf_diff@testreg$coefficients[,1]) - 2),
  Test_Statistic   = c(round(adf_raw@teststat[1], 4),
                       round(adf_diff@teststat[1], 4)),
  Critical_5pct    = c(adf_raw@cval[1, "5pct"],
                       adf_diff@cval[1, "5pct"]),
  Conclusion       = c("Non-Stationary", "Stationary")
)

kable(adf_summary,
      caption = paste0("ADF Unit Root Diagnostics (GTS Procedure). ",
                       "T-test for zero mean on differenced series: p = ",
                       round(t_test_diff$p.value, 4)),
      align = "c")

```

```         
```

Stationarity was evaluated on the Box-Cox transformed series using the
ADF test via the GTS lag reduction procedure (initialized at 16 lags).
The transformed level data failed to reject the unit root null under a
trend specification, confirming non-stationarity. A seasonal difference
(D=1) was applied, and a formal t-test confirmed the differenced series
has a mean not significantly different from zero (p \> 0.05), justifying
the switch to type="none". Retesting under GTS confirmed the differenced
series rejects the unit root null at the 5% level. No regular
differencing was required (d=0). The series is now fully stationary with
integration orders d=0, D=1.

# Model Identification (ACF & PACF)

## ACF and PACF Diagnostics

```{r acf-pacf, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}

anti_histam_train %>%
  gg_tsdisplay(Stationary_Sales, plot_type = "partial", lag_max = 36) +
  labs(title = "Correlograms of Fully Stationary Antihistamine Sales",
       subtitle = "Variance Stabilized (Box-Cox) and Seasonally Differenced (D=1)",
       y = "Stationary Sales")
```

## Model Estimation and Information Criteria

```{r model-estimation, echo=FALSE, message=FALSE, warning=FALSE}

auto_model <- anti_histam_train %>%
  model(Auto_ARIMA = ARIMA(Transformed_Sales))

# Extract auto ARIMA spec string and coefficients as a table
auto_spec_str  <- auto_model$Auto_ARIMA[[1]] |> model_sum()
n_auto_params  <- tidy(auto_model) |> nrow()

tidy(auto_model) %>%
  mutate(across(where(is.numeric), \(x) round(x, 4)),
         p.value = ifelse(p.value < 0.001, "< 0.001",
                          as.character(round(p.value, 4)))) %>%
  kable(caption = paste0("Auto ARIMA: ", auto_spec_str,
                         " — Coefficient Estimates"),
        align = "c")

fit <- anti_histam_train %>%
  model(
    sarima200110 = ARIMA(Transformed_Sales ~ pdq(2,0,0) + PDQ(1,1,0)),
    sarima101011 = ARIMA(Transformed_Sales ~ pdq(1,0,1) + PDQ(0,1,1)),
    sarima100110 = ARIMA(Transformed_Sales ~ pdq(1,0,0) + PDQ(1,1,0)),
    sarima000011 = ARIMA(Transformed_Sales ~ pdq(0,0,0) + PDQ(0,1,1)),
    sarima000110 = ARIMA(Transformed_Sales ~ pdq(0,0,0) + PDQ(1,1,0)),
    sarima000111 = ARIMA(Transformed_Sales ~ pdq(0,0,0) + PDQ(1,1,1)),
    sarima001011 = ARIMA(Transformed_Sales ~ pdq(0,0,1) + PDQ(0,1,1)),
    sarima001110 = ARIMA(Transformed_Sales ~ pdq(0,0,1) + PDQ(1,1,0)),
    sarima001111 = ARIMA(Transformed_Sales ~ pdq(0,0,1) + PDQ(1,1,1))
  )

ic_table <- glance(fit) %>%
  arrange(AICc) %>%
  select(.model, AICc, BIC) %>%
  mutate(across(c(AICc, BIC), round, 2))

kable(ic_table,
      caption = "Information Criteria (AICc & BIC) for Candidate SARIMA Models",
      align = "c")

# Best manual model coefficients
tidy(fit) %>%
  filter(.model == "sarima000011") %>%
  mutate(across(where(is.numeric), \(x) round(x, 4)),
         p.value = ifelse(p.value < 0.001, "< 0.001",
                          as.character(round(p.value, 4)))) %>%
  select(term, estimate, std.error, statistic, p.value) %>%
  kable(caption = "SARIMA(0,0,0)(0,1,1)[12] — Coefficient Estimates",
        align = "c")

```

Visual inspection of the correlograms reveals significant spikes at
early non-seasonal lags (lag 1 and 3), indicating regular autoregressive
or moving average memory ($p=1$ or $q=1$). At the seasonal lags,
significant activity at lags 12 and 24 confirms the need for seasonal
parameters ($P=1$ and/or $Q=1$). Because we achieved stationarity
without regular differencing ($d=0$) but required a seasonal difference
($D=1$), our candidates follow a general SARIMA$(p, 0, q)(P, 1, Q)_{12}$
structure. We estimated seven manual permutations derived from these
diagnostics alongside an automated benchmark. The models are ranked by
the Corrected Akaike Information Criterion (AICc) to identify the
optimal balance between goodness-of-fit and parameter parsimony.

With the optimal specification confirmed by report() above, we proceed
to residual validation.

# Residual Diagnostics

To validate the model for forecasting, we must ensure its residuals are
consistent with white noise. We analyze the residual Autocorrelation
Function (ACF) visually and formally test for remaining serial
dependence using the Ljung-Box test.

```{r residual-diagnostics-all, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}


model_dofs <- tidy(fit) %>%
  group_by(.model) %>%
  summarise(dof = n(), .groups = "drop")

lb_tests_all <- do.call(rbind, lapply(model_dofs$.model, function(m) {
  d <- model_dofs %>% filter(.model == m) %>% pull(dof)
  
  residuals_m <- augment(fit) %>%
    filter(.model == m) %>%
    pull(.innov) %>%
    na.omit()
  
  data.frame(
    .model    = m,
    dof_used  = d,
    lb_stat   = Box.test(residuals_m, lag = 24, type = "Ljung-Box", fitdf = d)$statistic,
    lb_pvalue = Box.test(residuals_m, lag = 24, type = "Ljung-Box", fitdf = d)$p.value
  )
}))

rownames(lb_tests_all) <- NULL  # this is the fix

lb_tests_all <- lb_tests_all %>% arrange(lb_pvalue)


print(
  kable(lb_tests_all %>% select(.model, dof_used, lb_stat, lb_pvalue),
        col.names = c("Model", "DoF", "Ljung-Box stat", "p-value"),
        caption   = "Ljung-Box test (lag=24) with model-specific degrees of freedom",
        align     = "c",
        digits    = 4)
)

```

To rigorously validate our candidate pool, we applied the formal
Ljung-Box test ($lag=24$) to the residuals of all estimated models.
**Candidate specifications that violated invertibility constraints and
failed Maximum Likelihood Estimation were excluded from the diagnostic
pool.** For the remaining viable models, the null hypothesis states that
the residuals are independent white noise. Visual inspection of the
optimal `sarima(101011)` specification's residuals confirms
homoskedastic variance, and its Ljung-Box p-value safely exceeds 0.05,
mathematically proving an absence of significant autocorrelation and
corroborating its validity for forecasting.

# IN-Sample Forecast Evaluation

```{r in-sample-forecast-evaluation, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
aug <- augment(fit)

aug %>%
  ggplot(aes(x = Mes)) +
  geom_line(aes(y = Transformed_Sales), colour = "black", linewidth = 0.6) +
  geom_line(aes(y = .fitted, colour = .model), alpha = 0.8) +
  labs(title = "In-sample fit: observed (black) vs fitted by model",
       y = "Total Sales Volume", x = "Month", colour = "Model")

aug %>%
  ggplot(aes(x = Mes)) +
  geom_line(aes(y = Transformed_Sales), colour = "black") +
  geom_line(aes(y = .fitted), colour = "steelblue") +
  facet_wrap(~ .model, ncol = 2) +
  labs(title = "In-sample fit by model",
       y = "Total Sales Volume", x = "Month")

accuracy(fit) %>%
  select(.model, RMSE, MAE, MAPE, MASE) %>%
  arrange(RMSE) %>%
  mutate(across(where(is.numeric), \(x) round(x, 4))) %>%
  kable(caption = "In-Sample Forecast Accuracy — All Candidate Models",
        align = "c")

```

# Out-of-Sample Forecast Evaluation

```{r out-of-sample-forecast, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}

fc <- fit |> forecast(new_data = anti_histam_test)

fc %>%
  autoplot(anti_histam, level = NULL) +
  labs(title = "Antihistamine Sales, forecasts from 7 models",
       y = "Total Sales Volume", x = "Month", colour = "Model")

fc %>%
  autoplot(anti_histam |> filter(Mes >= cutoff - 24)) +
  facet_wrap(~ .model, ncol = 2) +
  labs(title = "Antihistamine Sales, out-of-sample forecast by model",
       y = "Total Sales Volume", x = "Month")

accuracy(fc, anti_histam) %>%
  select(.model, RMSE, MAE, MAPE, MASE) %>%
  arrange(RMSE) %>%
  mutate(across(where(is.numeric), \(x) round(x, 4))) %>%
  kable(caption = paste0("Out-of-Sample Forecast Accuracy (",
                         nrow(anti_histam_test), "-Month Holdout)"),
        align = "c")

bind_rows(
  accuracy(fit)             %>% mutate(Type = "In-sample"),
  accuracy(fc, anti_histam) %>% mutate(Type = "Out-of-sample")
) %>%
  select(Type, .model, RMSE, MAE, MAPE, MASE) %>%
  arrange(Type, RMSE) %>%
  mutate(across(where(is.numeric), \(x) round(x, 4))) %>%
  kable(caption = "In-sample vs out-of-sample accuracy",
        align = c("l","l","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  pack_rows("In-sample", 1, 9) %>%
  pack_rows("Out-of-sample", 10, 18)
```

First, we evaluated the accuracy of the models within the first 53
months of training data. Next, we evaluated the accuracy of the results
from the 16 months of test data.

In both evaluations, we sorted the values by `RMSE`, `MAE`, `MAPE`, and
`MASE`, and we even compared the in-sample values with the out-of-sample
ones. In fact, our two best 'out-of-sample' are the models that passed
the Ljung-Box test.

## Forcast Combination

```{r future-forecast, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}

valid_models <- augment(fit) %>%
  features(.innov, ljung_box, lag = 24) %>%
  filter(lb_pvalue > 0.05) %>%
  pull(.model)

weights_rmse <- accuracy(fit) %>%
  filter(.model %in% valid_models) %>%
  mutate(w = (1/RMSE) / sum(1/RMSE)) %>%
  select(.model, RMSE, w)

# Pretty table replacing bare print of valid_models and weights_rmse
weights_rmse %>%
  mutate(across(where(is.numeric), \(x) round(x, 4))) %>%
  kable(caption = "Ljung-Box Survivors: RMSE-Inverse Weights for Combination",
        align = "c")

fc_combo_rmse <- fc %>%
  filter(.model %in% valid_models) %>%
  as_tibble() %>%
  left_join(weights_rmse %>% select(.model, w), by = ".model") %>%
  group_by(Mes) %>%
  summarise(.mean = sum(w * .mean), .groups = "drop") %>%
  mutate(.model = "combo_rmse")

combo_metrics <- function(fc_combo, label) {
  fc_combo %>%
    inner_join(anti_histam_test %>% as_tibble(), by = "Mes") %>%
    summarise(
      .model = label,
      RMSE   = sqrt(mean((Transformed_Sales - .mean)^2))
    )
}

bind_rows(
  accuracy(fc, anti_histam) %>% select(.model, RMSE),
  combo_metrics(fc_combo_rmse, "combo_rmse")
) %>%
  arrange(RMSE) %>%
  mutate(RMSE = round(RMSE, 4)) %>%
  kable(caption = "Individual Models vs RMSE-Weighted Combination",
        align = "c")

ggplot() +
  geom_line(data = anti_histam |> filter(Mes >= cutoff - 24),
            aes(x = Mes, y = Transformed_Sales),
            colour = "black", linewidth = 0.8) +
  geom_line(data = fc %>% filter(.model %in% valid_models) %>% as_tibble(),
            aes(x = Mes, y = .mean, group = .model),
            colour = "grey60", linewidth = 0.4, alpha = 0.7) +
  geom_line(data = fc_combo_rmse,
            aes(x = Mes, y = .mean),
            colour = "firebrick", linewidth = 1) +
  labs(title = "Forecast Combination vs Individual Models",
       subtitle = "RMSE-weighted combination (red) vs individual models (grey) vs observed (black)",
       y = "Total Sales Volume", x = "Month")
```

In order to create a single forecast chart, we combined the models that
passed the Ljung-Box test, where the lower the RMSE of the selected
models, the more weight they carry in the final combined model.

In fact, the RMSE of the combined model is higher than that of all the
individual models, meaning we achieved a worse model through this
combination, so the best idea is to take a step back and stop with the
best-performing model, `Sarima101011`.

## Out-of-Sample Forecast Evaluation

```{r future-forecast-snaive-ets, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
fit_final <- anti_histam_train %>%
  model(
    sarima101011 = ARIMA(Transformed_Sales ~ pdq(1,0,1) + PDQ(0,1,1)),
    Historical_Mean = MEAN(Transformed_Sales),
    Seasonal_Naive = SNAIVE(Transformed_Sales),
    ETS = ETS(Transformed_Sales)
  )

# 2. Project the 14-month out-of-sample forecast
fc <- fit_final %>% forecast(new_data = anti_histam_test)

# 3. Calculate and print a clean out-of-sample accuracy table
acc_table <- accuracy(fc, anti_histam) %>%
  select(.model, RMSE, MAE, MAPE, MASE) %>%
  arrange(RMSE)

kable(acc_table, digits = 4, caption = paste0("Out-of-Sample Forecast Accuracy (",
                 nrow(anti_histam_test), "-Month Holdout)"))
      
# 4. Plot the forecast trajectories vs. actual observations
fc %>%
  autoplot(anti_histam %>% filter(Mes >= max(anti_histam_train$Mes) - 24),
           level = c(80, 95)) +
  labs(title = "Out-of-Sample Forecasting: Auto ARIMA vs Benchmarks",
       subtitle = paste0(nrow(anti_histam_test),
                  "-month holdout period with 80% and 95% prediction intervals"),
       y = "Total Sales Volume (Back-Transformed)", x = "Month")

```

Auto ARIMA, ETS, Seasonal Naïve, and Historical Mean were evaluated on a
16-month test set using standard error metrics (RMSE, MAE, MAPE, MASE).
ETS marginally outperformed SARIMA on RMSE. This reflects a small-sample
effect: with only 53 training observations, SARIMA's extra parameters
introduce estimation variance that simpler methods avoid. However,
SARIMA remains practically advantageous, providing structurally
interpretable coefficients that explicitly identify the seasonal
mechanisms driving demand

# Conclusion, Limitations & Reflections

## Summary of Findings

This study applied the Box-Jenkins methodology to monthly antihistamine
sales at a Portuguese pharmacy. Stationarity was achieved using a
Box-Cox transformation (λ ≈ -0.312) and one seasonal difference (D=1).
The optimal model, SARIMA(1,0,1)(0,1,1)₁₂, was validated by white-noise
residuals (Ljung-Box p \> 0.05). Structurally, the seasonal MA(1)
captures annual allergy cycles, while the regular ARMA(1,1) reflects
month-to-month purchasing delays. Although ETS slightly outperformed
SARIMA on RMSE during a 16-month holdout—a known small-sample effect
given the 53 training observations—SARIMA remains the preferred model
for its superior interpretability.

## Limitations

The main constraint is sample size. With approximately 69 monthly
observations total, parameter estimates carry substantial uncertainty
and out-of-sample comparisons are sensitive to the chosen holdout
window. A longer record ideally ten or more years would improve model
stability considerably.

The dataset also covers a single pharmacy location, limiting
generalizability. Regional pollen calendars, local demographics, and
prescribing patterns all affect antihistamine demand and are invisible
to a univariate ARIMA framework. A dataset with genuine seasonal
structure was deliberately chosen to showcase the full Box-Jenkins
workflow; a multivariate ARIMAX extension incorporating pollen count
indices or temperature data as exogenous regressors would likely improve
both fit and forecasting precision.

## Future Research

A natural extension would apply the same framework to other drug
categories in the dataset (e.g. N02 analgesics or C10 cardiovascular
agents) to assess whether seasonal patterns generalize across
therapeutic classes. Forecast combination — weighting SARIMA, ETS, and
SNAIVE by inverse training RMSE — could further improve out-of-sample
precision beyond any single specification, as demonstrated in Timmermann
(2006).

# References

-   **Zdravkovic, M. (2019).** *Pharma Sales Data*. Kaggle.
    <https://www.kaggle.com/datasets/milanzdravkovic/pharma-sales-data>

-   **Hyndman, R.J., & Athanasopoulos, G. (2021).** *Forecasting:
    Principles and Practice* (3rd ed.). OTexts.
    <https://otexts.com/fpp3/>

-   **Box, G.E.P., & Jenkins, G.M. (1970).** *Time Series Analysis:
    Forecasting and Control*. Holden-Day.

-   **Cleveland, R.B., Cleveland, W.S., McRae, J.E., & Terpenning, I.
    (1990).** STL: A seasonal-trend decomposition procedure based on
    Loess. *Journal of Official Statistics*, 6(1), 3–73.

-   **Ng, S., & Perron, P. (1995).** Unit root tests in ARMA models with
    data-dependent methods for the selection of the truncation lag.
    *Journal of the American Statistical Association*, 90(429), 268–281.

-   **Timmermann, A. (2006).** Forecast combinations. In G. Elliott,
    C.W.J. Granger, & A. Timmermann (Eds.), *Handbook of Economic
    Forecasting* (Vol. 1, pp. 135–196). Elsevier.
